Appendix 2: Simulating 20-species landscapes

### 2015-11-25 Harris_2015/manuscript-materials/draft.tex
use the right version for the pdf and tex

\noindent \textbf{Recovering species interactions from simulated data.}
I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).
First, I used the rosalia package (Harris 2015a) to fit Markov network
models, as described above. For the analyses with 20 species, a weak
regularizer (equivalent to a logistic prior with scale 2) ensured that
the estimates were always finite.

@@

\noindent
\textbf{Twenty species.} In general, each model's performance was
highest for large landscapes with simple assembly rules and no
environmental heterogeneity (Figure 3). Despite some variability across
contexts, the rank ordering across methods was very consistent. In
particular, the four methods that controlled for indirect effects (the
Markov network, the generalized linear models, and the two partial
correlation-based methods) always matched or outperformed those that did
not. The Markov network consistently performed best of all. As
anticipated by Lee and Hastie (2012), generalized linear models closely
approximated the Markov network estimates (Figure 4A), especially when
the data sets were very large (Figure 3). As reviewed in Gotelli and
Ulrich (2009), however, most analyses in this field of ecology involve
fewer than 50 sites, and the gap between the Markov network and GLMs is
larger in this context. As shown in Appendix 4, the standard errors
associated with the estimates in Figure 3 are small (less than 0.01), so
the differences among methods should not be attributed to sampling
error.

@@

Finally, we can evaluate the models' statistical inferences (focusing on
the first two simulation types, for which the true interaction rates are
easiest to interpret). The Markov network's approximate Type I error
rate (defined here as the probability that 0 fell outside the 95\%
confidence interval for a pair of species where \(|\beta_{ij}|<0.1\))
depended on the simulation type: 0.02 for simulations that matched the
model's assumptions, versus 0.14 for simulations that included
environmental heterogeneity (see Appendix 4 for confidence interval
coverage across a range of \(\beta_{ij}\) values). The null model's Type
I error rates were 0.30 and 0.50 for the constant and heterogeneous
landscapes, respectively---far higher than the nominal 0.05 rate. Figure
4C shows, across a range of true interaction strengths, the probability
that the null model or the Markov network will predict the wrong sign of
the interaction with 95\% confidence. The null model makes such errors
more than 5 times as often as the Markov network, even though it only
reject the null hypothesis twice as often overall (Appendix 4). The
Markov network's errors were also more concentrated around 0, so it
never misclassified strong interactions like the null model did.

>>>>

\noindent \textbf{Recovering species interactions from simulated data.}
I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 4).
First, I used the rosalia package (Harris 2015a) to fit Markov network
models, as described above. For the analyses with 20 species, a weak
regularizer (equivalent to a logistic prior with scale 2) ensured that
the estimates were always finite.

@@

\noindent
\textbf{Twenty species.} In general, each model's performance was
highest for large landscapes with simple assembly rules and no
environmental heterogeneity (Figure 3). Despite some variability across
contexts, the rank ordering across methods was very consistent. In
particular, the four methods that controlled for indirect effects (the
Markov network, the generalized linear models, and the two partial
correlation-based methods) always matched or outperformed those that did
not. The Markov network consistently performed best of all. As
anticipated by Lee and Hastie (2012), generalized linear models closely
approximated the Markov network estimates (Figure 4A), especially when
the data sets were very large (Figure 3). As reviewed in Gotelli and
Ulrich (2009), however, most analyses in this field of ecology involve
fewer than 50 sites, and the gap between the Markov network and GLMs is
larger in this context. As shown in Appendix 5, the standard errors
associated with the estimates in Figure 3 are small (less than 0.01), so
the differences among methods should not be attributed to sampling
error.

@@

Finally, we can evaluate the models' statistical inferences (focusing on
the first two simulation types, for which the true interaction rates are
easiest to interpret). The Markov network's approximate Type I error
rate (defined here as the probability that 0 fell outside the 95\%
confidence interval for a pair of species where \(|\beta_{ij}|<0.1\))
depended on the simulation type: 0.02 for simulations that matched the
model's assumptions, versus 0.14 for simulations that included
environmental heterogeneity (see Appendix 5 for confidence interval
coverage across a range of \(\beta_{ij}\) values). The null model's Type
I error rates were 0.30 and 0.50 for the constant and heterogeneous
landscapes, respectively---far higher than the nominal 0.05 rate. Figure
4C shows, across a range of true interaction strengths, the probability
that the null model or the Markov network will predict the wrong sign of
the interaction with 95\% confidence. The null model makes such errors
more than 5 times as often as the Markov network, even though it only
reject the null hypothesis twice as often overall (Appendix 5). The
Markov network's errors were also more concentrated around 0, so it
never misclassified strong interactions like the null model did.
